### 2019-05-08 papers/haskell19/1-intro.tex
Minor changes, fixed layout, added citation

% --------------------------------------
% Intro
%
Random property-based testing is a powerful technique for finding bugs
\cite{ArtsHNS15, HughesPAN16, HughesNSA16, MidtgaardJKNN17}.
%
In the Haskell community, QuickCheck is the predominant tool for that task
\cite{ClaessenH00}.
%
The developers specify
%
\begin{inparaenum}[(i)]
\item the testing properties their systems must fulfill, and
\item random data generators (or generators for short) for the data types
  involved at their properties.
\end{inparaenum}
%
Then, QuickCheck generates random values using these generators, which are used
to evaluate the system's properties in search of possible counterexamples, which
often indicate the presence of bugs.

@@

\begin{code}
data Html  =  Text  String
           |  Sing  String
           |  Tag   String Html
           |  Join  Html Html
\end{code}
%
This type allows us to build HTML pages via four possible data constructors:
|Text| is used for plain text values; |Sing| and |Tag| represent singular and
paired HTML tags, respectively; and |Join| simply concatenates two HTML pages
one after another.
%
Note that the constructors |Tag| and |Join| are recursive, as they have at least
one field of their own type.
%
In this paper, we will refer to any non-recursive data constructor as
\emph{terminal}, and \emph{non-terminal} in any other case.

@@

% --------------------------------------
% Automatic derivation tools
%
As one can observe, the previous definition is quite mechanical, and depends
only on the generation frequencies we choose for each constructor.
%
Fortunately, there exists a variety of tools for obtaining random generators
automatically, based on the information encoded on the definition of the types
we want to generate.
%
Most of them implemented using Haskell's meta-programming framework named
Template Haskell \cite{SheardJ02}.
%
For instance, tools like \emph{MegaDeTH} \cite{GriecoCB16, grieco2017} or
\emph{DRAGEN} \cite{DBLP:conf/haskell/MistaRH18} will synthesize random
generators using roughly the same procedure as the one shown above.
%
On one hand, \emph{MegaDeTH} derives generators which simply pick among data
constructors with uniform probabilities, whereas \emph{DRAGEN} analyzes type
definitions, tuning the generation frequencies based on a predictive mechanism.
%
This allows the users to specify the desired distribution of values of the
derived generators.
%
\looseness=-1

@@

%
% Contributions
%
The main contribution of this paper are:
%
\begin{CompactItemize}
\item We identify some important limitations of performing random testing with
  automatically derived generators using state-of-the-art tools (Section
  \ref{sec:sources}).
\item We present an extensible mechanism for representing random values built
  upon different sources of structural information using Data Types \`a la Carte
  (Section \ref{sec:representation}).
\item We develop a modular generation schema, exploiting the composable nature
  of our representations, which are extended to encode information relevant to
  the generation process at the type level (Section \ref{sec:generators}).
\item We propose a simple type-level idiom for describing extensible generators,
  based on the types used to represent the desired shape of our random data
  (Section \ref{sec:typelevel}).\looseness=-1
\item We provide a Template Haskell tool%
  %
  \footnote{Link will be made available for the final version of this paper.}
  %
  for automatically deriving all the required machinery presented throughout
  this paper, and evaluate its generation performance using three real-world
  case studies (Section \ref{sec:casestudies}).
\end{CompactItemize}
%
Overall, we present a novel technique for reusing automatically derived
generators in a composable fashion, in contrast with the usual paradigm of
automatically deriving ``rigid'', monolithic generators.

>>>>

% --------------------------------------
% Intro
%
Random property-based testing is a powerful technique for finding bugs
\cite{ArtsHNS15, HughesPAN16, HughesNSA16, MidtgaardJKNN17}.
%
In the Haskell community, QuickCheck is the predominant tool for that task
\cite{ClaessenH00}.
%
The developers specify
%
\begin{inparaenum}[(i)]
\item the testing properties their systems must fulfill, and
\item random data generators (or generators for short) for the data types
  involved at their properties.
\end{inparaenum}
%
Then, QuickCheck generates random values using these generators, and uses them
to evaluate the testing properties in search of possible counterexamples, which
often indicate the presence of bugs.

@@

\begin{code}
data Html  =  Text  String
           |  Sing  String
           |  Tag   String Html
           |  Join  Html Html
\end{code}
%
This type allows us to build HTML pages via four possible data constructors:
|Text| is used for plain text values; |Sing| and |Tag| represent singular and
paired HTML tags, respectively; and |Join| simply concatenates two HTML pages
one after another.
%
Note that the constructors |Tag| and |Join| are recursive, as they have at least
one field of their own type.
%
In this paper, we will refer to any recursive data constructor as
\emph{non-terminal}, and as \emph{terminal} in any other case.

@@

% --------------------------------------
% Automatic derivation tools
%
As one can observe, the previous definition is quite mechanical, and depends
only on the generation frequencies we choose for each constructor.
%
Fortunately, there exists a variety of tools for obtaining random generators
automatically, based on the information encoded on the definition of the types
we want to generate.
%
Most of them implemented using Haskell's meta-programming framework named
Template Haskell \cite{SheardJ02}.
%
For instance, tools like \emph{MegaDeTH} \cite{GriecoCB16, grieco2017} or
\emph{DRAGEN} \cite{DBLP:conf/haskell/MistaRH18} will synthesize random
generators using roughly the same generation procedure as the one shown above.
%
On one hand, \emph{MegaDeTH} derives generators which simply pick among data
constructors with uniform probabilities, whereas \emph{DRAGEN} analyzes type
definitions, tuning the generation frequencies based on a predictive mechanism.
%
This allows the users to specify the desired distribution of values produced by
the derived generators.
%
\looseness=-1

@@

%
% Contributions
%
The main contribution of this paper are:
%
\begin{CompactItemize}
\item We identify some important limitations of performing random testing with
  automatically derived generators using state-of-the-art tools (Section
  \ref{sec:sources}).
\item We present an extensible mechanism for representing random values built
  upon different sources of structural information using Data Types \`a la Carte
  (Section \ref{sec:representation}).
\item We develop a modular generation schema, exploiting the composable nature
  of our representations, which are extended to encode information relevant to
  the generation process at the type level (Section \ref{sec:generators}).
\item We propose a simple type-level idiom for describing extensible generators,
  based on the types used to represent the desired shape of our random data
  (Section \ref{sec:typelevel}).\looseness=-1
\item We provide a Template Haskell tool%
  %
  \footnote{Link will be made available for the final version of this paper.}
  %
  for automatically deriving all the required machinery presented throughout
  this paper, and evaluate its generation performance using three real-world
  case studies (Section \ref{sec:casestudies}).
\end{CompactItemize}
%
Overall, we present a novel technique for reusing automatically derived
generators in a composable fashion, in contrast to the usual paradigm of
automatically deriving ``rigid'', monolithic generators.

### 2019-05-08 papers/haskell19/2-sources.tex
Minor changes, fixed layout, added citation

(<+>) :: Html -> Html -> Html
(<+>) x y = Join x (Join br y)
\end{code} %$
%
These functions let us encode structured HTML constructions like line-breaks
(|br|), bold blocks (|bold|), unordered lists (|list|) and concatenation of
values one below another (|<+>|).
%
\looseness=-1

@@

In our tests, we were not able to exercise any of these two patterns more than
6\% of the overall testing time (and considering recursive sub-calls), using
random generators derived using either \emph{MegaDeTH} and \emph{DRAGEN}.
%
As expected, most of the random test cases were exercising the simplest (an
rather uninteresting) patterns of this function, corresponding to its last three
clauses.

>>>>

(<+>) :: Html -> Html -> Html
(<+>) x y = x iJoin br iJoin y
\end{code} %$
%
These functions let us encode structured HTML constructions like line-breaks
(|br|), bold blocks (|bold|), unordered lists (|list|) and concatenation of
values one below another (|<+>|).
%
\looseness=-1

@@

In our tests, we were not able to exercise any of these two patterns more than
6\% of the overall testing time (and considering recursive sub-calls), using
random generators derived using both \emph{MegaDeTH} and \emph{DRAGEN}.
%
As expected, most of the random test cases were exercising the simplest (an
rather uninteresting) patterns of this function, corresponding to its last three
clauses.

### 2019-05-08 papers/haskell19/3-representation.tex
Minor changes, fixed layout, added citation

% --------------------------------------
% The basic idea
%
This section introduces a unified representation for the different constructions
we might want to consider when generating random values.
%
The key idea of this work is to lift each different source of structural
information at the type level.
%
In this light, the ``shape'' of our random data is determined entirely by the
type we use to represent it during the generation process.

For this, we will define a set of different simple ``open'' types, each one in
charge of representing a single construction of our \emph{target} data type,
i.e., the actual data type we want to randomly generate.
%
These types can be later
%
\begin{inparaenum}[(i)]
\item combined in several ways depending on the desired shape of our test data
  (applying the familiar \`a la Carte technique);
\item randomly generated (see Section \ref{sec:generators}); and finally,
\item transformed back to the corresponding values of the target data type
  automatically.
\end{inparaenum}
%
This lifted representation of constructions can be automatically derived from
our source code in compile-time, relieving programmers of the burden of manually
implementing the required machinery.

@@

Note how every definition regarding each different constructor so far was
obtained rather mechanically.
%
This let us automate the process quite a bit using Haskell's meta-programming
capabilities.

@@

So far we have seen how to represent each data constructor of our |Html| data
type independently.
%
In order to represent interesting values, we need to be able to combine these
single representations into (possibly complex) composite ones.
%
For this purpose, we will define a generic functor type |oplus| to encode the
choice between two given representation types as follows:

@@

\end{code}
%
This infix type-level operator let us combine two representations |f| and |g|
into a composite one |f oplus'' g|, representing either a value from |f| (via
the |InL| constructor) or a value from |g| (via the |InR| constructor).
%
This operator works pretty much in the same way as Haskell's |Either| data type,
except that, instead of combining two base types, it works combining two
\emph{parametric type constructors}, hence the kind signature |* -> *| in both
|f| and |g|.
%
For instance, we can use the type |Con_Text oplus'' Con_Tag| to encode values
representing either plain text HTMLs or paired tags.
%
Such values can be constructed easily using the injections |InL| and |InR| on
each each case, respectively.

@@

\begin{code}
instance (Algebra f a, Algebra g a) => Algebra (f oplus'' g) a
  where  alg (InL fa) = alg fa
         alg (InR ga) = alg ga
\end{code}
%
where we simply use the appropriate |Algebra| instance for the inner
representation, based on which injection was used to construct the composite
value.
%
It is worth remarking that multiple compositions are handled transparently,
i.e., for every set of types |f1|, |f2|, ..., |fk| representing a target type
|t| and having suitable |Algebra| type instances, then the composition |f1
oplus'' f2 oplus'' dots oplus'' fk| can be mapped back to |t| using this same
mechanism.
%
Additionally, the order in which we associate each operand of |oplus| results
semantically irrelevant, thus we can omit parentheses when describing composite
representations.
%
However, in practice, the way in which we associate our constructions affects
the generation speed dramatically.
%
This phenomena is addressed in detail in Section \ref{sec:casestudies}.
%
For simplicity here, we will assume |oplus| as having right associativity, i.e.,
|f oplus'' g oplus'' h == f oplus'' (g oplus'' h)|.

@@

\begin{code}
newtype Fix (f :: * -> *) = FixC { unFix :: f (Fix f) }
\end{code}
%
Given a parametric type |f| of kind |* -> *| representing some target data type,
the type |Fix f| instantiates each recursive field of |f| with |Fix f|, thus
closing the definition of |f| into itself---note thus that the kind of |Fix f|
results |*|.
%
In addition, note that we defined |Fix| as a |newtype|.
%
This way, the compiler will avoid producing a runtime representation for it,
helping to reduce some of the unnecessary overhead.
%
This is possible since we only use it as a wrapper to obtain a self-recursive
type from a parametric one.

@@

The effect of a fixed point combinator is much easier to interpret considering
an example.
%
Let us imagine we want to represent our |Html| data type using all of its data
constructors.
%
For this purpose we use the following type to represent it:

@@

Note that, by representing values obtained from function application this way,
we are not performing any actual computation---we simply store the functions'
input arguments.
%
Instead, the actual functions are evaluated when transforming our
representations back to their target type, by the means of an |Algebra|:

@@

It is important to remark that this approach inherits any possible downside from
the functions we use to represent our target data.
%
In particular, using non-terminating functions to represent our data might
produce a non-terminating behavior when calling to the |eval| function.

@@

\begin{code}
simplify (Join (Text t1) (Text t2))   = dots
simplify (Join (Join (Text t1) x) y)  = dots
\end{code}
%
We can extend our approach in order to represent these patterns as well.
%
We start defining single data types for each one of them, this time encoding in
the fields of each single data constructor the free pattern variables (or
wildcards) appearing in its corresponding pattern, from left to right:

@@

Using the constructions presented so far, we are now able to represent values of
our target data using a higher level of abstraction.
%
Instead of constructing values solely by combining different data constructors,
this mechanism lets us also consider complex constructions like abstract
interface functions and functions' patterns, treating them uniformingly and in a
fine grained fashion.

As we have introduced in the Section \ref{sec:sources}, having an extensible,
fine-grained framework for generating random data might be of great advantage,
as we would like to generate random test cases following different strategies
when testing different properties or sub-systems.
%
For instance, generating values satisfying the pattern matchings of the function
|simplify| could be relevant only when testing properties involving such
function.
%
For this purpose, we could represent |Html| values as follows:
%
\begin{code}
type Html_simplify  =      Con_Text oplus'' Con_Sing oplus'' Con_Tag oplus'' Con_Join
                    oplus  Pat_simplify_1 oplus'' Pat_simplify_2
\end{code}

>>>>

% --------------------------------------
% The basic idea
%
This section introduces a unified representation for the different constructions
we might want to consider when generating random values.
%
The key idea of this work is to lift each different source of structural
information to the type level.
%
In this light, the ``shape'' of our random data is determined entirely by the
type we use to ``represent'' it during the generation process.

For this purpose, we will define a set of different simple ``open'' types, each
one in charge of representing a single random construction from our
\emph{target} data type, i.e., the actual data type we want to randomly
generate.
%
These types can be later
%
\begin{inparaenum}[(i)]
\item combined in several ways depending on the desired shape of our test data
  (applying the familiar \`a la Carte technique);
\item randomly generated (see Section \ref{sec:generators}); and finally,
\item transformed back to the corresponding values of the target data type
  automatically.
\end{inparaenum}
%
This lifted representation of constructions can be automatically derived from
our source code in compile-time, relieving programmers of the burden of manually
implementing the required machinery.

@@

Note how every definition regarding each different constructor so far was
obtained rather mechanically.
%
This let us automate the process quite a bit using Template Haskell.

@@

So far we have seen how to represent each data constructor of our |Html| data
type independently.
%
In order to represent interesting values, we need to be able to combine these
single representations into (possibly complex) composite ones.
%
For this purpose, we will define a generic functor type |oplus| to encode the
choice between two given representation types:

@@

\end{code}
%
This infix type-level operator let us combine two representations |f| and |g|
into a composite one |f oplus'' g|, representing either a value from |f| (via
the |InL| constructor) or a value from |g| (via the |InR| constructor).
%
This operator works pretty much in the same way as Haskell's |Either| data type,
except that, instead of combining two base types, it works combining two
\emph{parametric type constructors}, hence the kind signature |* -> *| in both
|f| and |g|.
%
For instance, we can use the type |Con_Text oplus'' Con_Tag| to encode values
representing either plain text HTMLs or paired tags.
%
Such values can be constructed using the injections |InL| and |InR| on each
case, respectively.
%
\looseness=-1

@@

\begin{code}
instance (Algebra f a, Algebra g a) => Algebra (f oplus'' g) a
  where  alg (InL fa) = alg fa
         alg (InR ga) = alg ga
\end{code}
%
where we simply use the appropriate |Algebra| instance for the inner
representation, based on which injection was used to construct the composite
value.
%
It is worth remarking that multiple compositions are handled transparently,
i.e., for every set of types |f1|, |f2|, ..., |fk| representing a target type
|t| and having suitable |Algebra| type instances, then the composition |f1
oplus'' f2 oplus'' dots oplus'' fk| can be mapped back to |t| using this same
mechanism.
%
Additionally, the order in which we associate each operand of |oplus| results
semantically irrelevant, thus we can omit parentheses when describing composite
representations.
%
However, in practice, the way in which we associate our constructions affects
the generation speed dramatically.
%
This phenomenon is addressed in detail in Section \ref{sec:casestudies}.
%
For simplicity here, we will assume |oplus| as having right associativity, i.e.,
|f oplus'' g oplus'' h == f oplus'' (g oplus'' h)|.

@@

\begin{code}
newtype Fix (f :: * -> *) = FixC { unFix :: f (Fix f) }
\end{code}
%
Given a parametric type |f| of kind |* -> *| representing some target data type,
the type |Fix f| instantiates each recursive field of |f| with |Fix f|, thus
closing the definition of |f| into itself---thus the kind of |Fix f| results
|*|.
%
In addition, note that we defined |Fix| as a |newtype|.
%
This way, the compiler will avoid producing a runtime representation for it,
helping to reduce some of the unnecessary overhead.
%
% This is possible since we only use it as a wrapper to obtain a self-recursive
% type from a parametric one.

@@

The effect of a fixed point combinator is much easier to interpret considering
an example.
%
Let us imagine we want to represent our |Html| data type using all of its data
constructors.
%
For this purpose, we can use the following type:

@@

Note that, by representing values obtained from function application this way,
we are not performing any actual computation---we simply store the functions'
input arguments.
%
Instead, these functions are evaluated when transforming our representations
back to their target type, by the means of an |Algebra|:

@@

It is important to remark that this approach inherits any possible downside from
the functions we use to represent our target data.
%
In particular, representing non-terminating functions might produce a
non-terminating behavior when calling to the |eval| function.

@@

\begin{code}
simplify (Join (Text t1) (Text t2))   = dots
simplify (Join (Join (Text t1) x) y)  = dots
\end{code}
%
We can extend our approach in order to represent these patterns as well.
%
We start by defining data types for each one of them, this time using the fields
of each single data constructor to encode the free pattern variables (or
wildcards) appearing on its corresponding pattern, from left to right:

@@

Using the constructions presented so far, we are now able to represent values of
our target data using a higher level of abstraction.
%
Instead of constructing values solely by combining data constructors, this
mechanism lets us also consider complex constructions like abstract interface
functions and functions' patterns, treating them uniformingly and in a fine
grained fashion.

As we have introduced in the Section \ref{sec:sources}, having an extensible
framework for generating random data might be of great advantage, as we would
like to generate random test cases following different strategies when testing
different properties or sub-systems.
%
For instance, generating values satisfying the pattern matchings of the function
|simplify| could be relevant only when testing properties involving such
function.
%
For this purpose, we could represent |Html| values as follows:
%
\begin{code}
type Html_simplify  =      Con_Text oplus'' Con_Sing oplus'' Con_Tag oplus'' Con_Join
                    oplus  Pat_simplify_1 oplus'' Pat_simplify_2
\end{code}

### 2019-05-08 papers/haskell19/4-generators.tex
NIT
Some compresUsing CompactItemize and arranged code to not break on different pages
Minor changes, fixed layout, added citation

% --------------------------------------
% Intro
%
So far we have seen how to encode different constructions representing
interesting values from our target data types.
%
Such representations follow a modular approach, where each different
construction is considered independent from the rest.
%
This modularity allows us to derive each different construction independently,
as well to specify the shape of our target data with a fine granularity.

In this section, we introduce the machinery required to randomly generate the
values encoded using our representations.
%
This step also follows the modular, fine-grained fashion, resulting in a
generation process entirely compositional.
%
In this light, our random generators are built from smaller ones (each one
representing a single random construction), and are solely based on the types we
use to represent the shape of our random data.

Ideally, our aim is to be able to obtain random generators with a behavior
similar to the one presented for |Html| in Section \ref{sec:intro}.
%
If we take a closer look at its definition, we can observe three factors
happening simultaneously:
%
\begin{itemize}
\item We use QuickCheck's generation size to limit the depth of the generated
  values, reducing it by one on each recursive call of the local auxiliary
  function |gen|.
\item We differentiate between \emph{terminal} and \emph{non-terminal
    constructors}, limiting our generator to pick only among terminal ones when
  we have reached the maximum depth (case |gen 0|).
\item We generate different constructions in a (possibly) different frequency.
\end{itemize}
%
For the rest of this section, we will focus on modeling these aspects in our
modular framework, in such a way that does not compromise the compositionality
obtained so far.

@@

\begin{code}
class BArbitrary1 (f :: * -> *) where
  liftBGen :: BGen a -> BGen (f a)
\end{code}
%
Note the similarities between the |Arbitrary1| and |BArbitrary1| type classes.
%
We will use this type class to implement random generators for each construction
we are considering.
%
Recalling our |Html| example, we can define modular random generators for the
constructions representing data constructors as follows:

@@

As we have shown so far, generating each representation is rather
straightforward: there is only one data constructor to select, and every field
is generated following a mechanical recipe.
%
In our approach, most of the generation complexity is encoded in the random
generator used for composite representations, which are built upon the |oplus|
operator.
%
Before introducing it, we need to define some additional machinery to encode the
notions of terminal construction and generation frequency.

@@

On the other hand, the last step of our generation process, which only generates
terminal constructions, could be seen as considering the non-terminal ones as
having generation frequency zero.
%
This way, we will introduce another type-level computation to calculate the
\emph{terminal generation frequency} |FreqOf'| of a given representation as
follows:

@@

\begin{code}
instance (BArbitrary1 f, BArbitrary1 g)
  => BArbitrary1 (f oplus g) where
  liftBGen bgen d =
    if d > 0
    then frequency
      [ (freqVal  at(FreqOf f) ,  InL <$> liftBGen bgen d)
      , (freqVal  at(FreqOf g) ,  InR <$> liftBGen bgen d) ]
    else frequency
      [ (freqVal  at(FreqOf' f) ,  InL <$> liftBGen bgen d)
      , (freqVal  at(FreqOf' g) ,  InR <$> liftBGen bgen d) ]
\end{code}
%
Like the generator for |Html| introduced in Section \ref{sec:intro}, this
generator branches over the current generation depth |d|.
%
In the case we can still generate values from any construction (|d > 0|), we
will use QuickCheck's |frequency| operation to randomly choose between
generating a value of each side of the |oplus|, i.e., either a value of |f| or a
value of |g|, following the generation frequencies specified for both of them,
and wrapping the values with the appropriate injection |InL| or |InR| on each
case.
%
Such frequencies are obtained by \emph{reflecting} the type-level natural values
obtained from applying |FreqOf| to both |f| and |g|, using a type-dependent
function |freqVal| that returns the number corresponding to the type-level
natural value we apply to it:

\begin{code}
freqVal :: forall n dot KnownNat n => Int
\end{code}
%
Note that the type of |freqVal| is ambiguous, since it quantifies over every
possible known type-level natural value |n|.
%
We use a \emph{visible type application} (employing the |at (...)| syntax) to
disambiguate to which natural value we are actually referring to.
%
Then, for instance, the value |freqVal at (FreqOf (f otimes'' 5))| will evaluate
to |5 :: Int|.

The rest of our random generator works analogously, except that, this time we
only want to generate terminal constructions, hence we use the |FreqOf'| type
family to compute the terminal generation frequency of each operand.
%
If any of |FreqOf' f| or |FreqOf' g| evaluates to zero, then it means that that
operand does not contain any terminal constructions, and |frequency| will not
consider it when generating terminal values.

If it happens that both |FreqOf' f| and |FreqOf' g| compute to zero when
generating values from composite representation, then it will produce a runtime
error triggered by the function |frequency|, as it does not have anything with a
positive frequency to generate.
%
This kind of exceptions will arise, for example, if we forget to tag at least a
one construction from our final representation as terminal.
%
Fortunately, such runtime exceptions can be catched at compile-time.
%
We can define a type constraint |HasTerminals| that ensures we are trying to
generate values using a representation with a strictly positive terminal
generation frequency---thus containing at least a single terminal construction:

@@

\begin{code}
genRep  :: forall f a dot  ( BArbitrary1 f
                           , HasTerminals f
                           , Algebra f a) => BGen a
genRep d = eval <$> bounded at (Fix f) d
\end{code} %$
%
Note how this primitive is also ambiguous in the type used for the
representation.
%
This allows us to use a visible type application to obtain values from the same
target type but generated using a different underlying representations, and
avoiding the use of |Proxy| values everywhere.
%
For instance, we can obtain two different concrete generators for our |Html|
type simply by changing its representation type as follows:

>>>>

% --------------------------------------
% Intro
%
So far we have seen how to encode different random constructions representing
interesting values from our target data types.
%
Such representations follow a modular approach, where each construction is
independent from the rest.
%
This modularity allows us to derive each different construction individually, as
well to specify the shape of our target data with a fine granularity.

In this section, we introduce the machinery required to randomly generate the
values encoded using our representations.
%
This step also follows the modular fashion, resulting in a generation process
entirely compositional.
%
In this light, our random generators are built from smaller ones (each one
representing a single random construction), and are solely based on the types we
use to represent the shape of our random data.

Ideally, our aim is to be able to obtain random generators with a behavior
similar to the one presented for |Html| in Section \ref{sec:intro}.
%
If we take a closer look at its definition, we can observe three factors
happening simultaneously:
%
\vspace{2pt}
\begin{CompactItemize}
\item We use QuickCheck's generation size to limit the depth of the generated
  values, reducing it by one on each recursive call of the local auxiliary
  function |gen|.
\item We differentiate between \emph{terminal} and \emph{non-terminal
    constructors}, picking only among terminal ones when we have reached the
  maximum depth (case |gen 0|).
\item We generate different constructions in a (possibly) different frequency.
\end{CompactItemize}
%
For the rest of this section, we will focus on modeling these aspects in our
modular framework, in such a way that does not compromise the compositionality
obtained so far.

@@

\begin{code}
class BArbitrary1 (f :: * -> *) where
  liftBGen :: BGen a -> BGen (f a)
\end{code}
%
Note the similarities between |Arbitrary1| and |BArbitrary1|.
%
We will use this type class to implement random generators for each construction
we are considering.
%
Recalling our |Html| example, we can define modular random generators for the
constructions representing data constructors as follows:

@@

As we have seen so far, generating each representation is rather
straightforward: there is only one data constructor to pick, and every field is
generated using a mechanical recipe.
%
In our approach, most of the generation complexity is encoded in the random
generator for composite representations, built upon the |oplus| operator.
%
Before introducing it, we need to define some additional machinery to encode the
notions of terminal construction and generation frequency.
%
\looseness=-1

@@

Furthermore, the last step of our generation process, which only generates
terminal constructions, could be seen as considering the non-terminal ones as
having generation frequency zero.
%
This way, we can introduce another type-level computation to calculate the
\emph{terminal generation frequency} |FreqOf'| of a given representation as
follows:

@@

\begin{code}
instance (BArbitrary1 f, BArbitrary1 g)
  => BArbitrary1 (f oplus g) where
  liftBGen bgen d =
    if d > 0
    then frequency
      [ (freqVal  at(FreqOf f) ,  InL <$> liftBGen bgen d)
      , (freqVal  at(FreqOf g) ,  InR <$> liftBGen bgen d) ]
    else frequency
      [ (freqVal  at(FreqOf' f) ,  InL <$> liftBGen bgen d)
      , (freqVal  at(FreqOf' g) ,  InR <$> liftBGen bgen d) ]
\end{code}
%
Like the generator for |Html| introduced in Section \ref{sec:intro}, this
generator branches over the current depth |d|.
%
In the case we can still generate values from any construction (|d > 0|), we
will use QuickCheck's |frequency| operation to randomly choose between
generating a value of each side of the |oplus|, i.e., either a value of |f| or a
value of |g|, following the generation frequencies specified for both of them,
and wrapping the values with the appropriate injection |InL| or |InR| on each
case.
%
Such frequencies are obtained by \emph{reflecting} the type-level natural values
obtained from applying |FreqOf| to both |f| and |g|, using a type-dependent
function |freqVal| that returns the number corresponding to the type-level
natural value we apply to it:

\begin{code}
freqVal :: forall n dot KnownNat n => Int
\end{code}
%
Note that the type of |freqVal| is ambiguous, since it quantifies over every
possible known type-level natural value |n|.
%
We use a \emph{visible type application} \cite{eisenberg2016visible} (employing
the |at (...)| syntax) to disambiguate to which natural value we are actually
referring to.
%
Then, for instance, the value:

\begin{code}
  freqVal at (FreqOf (f otimes'' 5 oplus' g otimes' 4))
\end{code}
%
evaluates to the concrete value |9 :: Int|.

The |else| clause of our random generator works analogously, except that, this
time we only want to generate terminal constructions, hence we use the |FreqOf'|
type family to compute the terminal generation frequency of each operand.
%
If any of |FreqOf' f| or |FreqOf' g| evaluates to zero, it means that such
operand does not contain any terminal constructions, and |frequency| will not
consider it when generating terminal values.

Moreover, if it happens that both |FreqOf' f| and |FreqOf' g| compute to zero
simultaneously, then this will produce a runtime error triggered by the function
|frequency|, as it does not have anything with a positive frequency to generate.
%
This kind of exceptions will arise, for example, if we forget to tag at least a
one construction from our final representation as terminal.
%
Fortunately, such runtime exceptions can be catched at compile-time.
%
We can define a type constraint |HasTerminals| that ensures we are trying to
generate values using a representation with a strictly positive terminal
generation frequency---thus containing at least a single terminal construction:

@@

\begin{code}
genRep  :: forall f a dot  ( BArbitrary1 f
                           , HasTerminals f
                           , Algebra f a) => BGen a
genRep d = eval <$> barbitrary at (Fix f) d
\end{code} %$
%
Note how this primitive is also ambiguous in the type used for the
representation.
%
This allows us to use a visible type application to obtain values from the same
target type but generated using different underlying representations, and
avoiding the use of |Proxy| values everywhere.
%
For instance, we can obtain two different concrete generators for our |Html|
type simply by changing its representation type as follows:

### 2019-05-08 papers/haskell19/5-typelevel.tex
Minor changes, fixed layout, added citation

% --------------------------------------
% Naming conventions
%
% To begin with, let us consider the name conventions we used when defining our
% representation data types along this paper.
%
% Since our aim is to automate this process as much as possible using Haskell's
% meta-programming capabilities,
%
The first problem we face is that of naming conventions.
%
In practice, the actual names used when deriving construction representations
need to be generated in such a way that they are compliant with Haskell's
syntax, and also that they are \emph{unique within our namespace}.
%
This means that, type names like |Fun_cat| or |Pat_simplify_1| are, technically,
not valid Haskell data type names, thus they will have to be synthesized as
something like |Fun_lt_plus_gt_543| and |Pat_simplify_1_325|, where the last
sequence of numbers is inserted by Template Haskell to ensure uniqueness.

@@

\begin{code}
type family Con (c :: Symbol)             :: * -> *
type family Fun (f :: Symbol)             :: * -> *
type family Pat (p :: Symbol) (n :: Nat)  :: * -> *
\end{code}
%
where |Symbol| is the kind of type-level strings in Haskell.
%
Then, our derivation process will synthesize each representation using unique
names, along with a type instance of the corresponding type family.
%
For instance, along with the representations of |Con_Text|, |Fun_cat| and
|Pat_simplify_1|, we will automatically derive the following type instances:

@@

% The biggest different appears when deriving the required type instances for the
% |Con|, |Fun| and |Pat| synonyms.
%
However, instead of carrying this type parameter in our generation
specifications, we can avoid it by hiding it using an existential type:

@@

\begin{code}
genIntTree :: BGen (Tree Int)
genIntTree = genRep at (Tree_Spec apply' Int)
\end{code}
%
Where |apply| is a closed type family that traverses our generation
specification, applying the |Int| type to every occurrence of |Some|:

@@

Along with our automated constructions derivation mechanism, the machinery
introduced in this section allows us to specify random generators using a
concise and flexible type-level combinators language.
%
The next section evaluates our approach in terms of performance using a set of
case studies extracted from real-world Haskell implementations.

>>>>

% --------------------------------------
% Naming conventions
%
% To begin with, let us consider the name conventions we used when defining our
% representation data types along this paper.
%
% Since our aim is to automate this process as much as possible using Haskell's
% meta-programming capabilities,
%
The first problem we face is that of naming conventions.
%
In practice, the actual name used when deriving the representation for each
random construction needs to be generated in such a way that is complies with
Haskell's syntax, and also that it is \emph{unique within our namespace}.
%
This means that, type names like |Fun_cat| or |Pat_simplify_1| are, technically,
not valid Haskell data type names, thus they will have to be synthesized as
something like |Fun_lt_plus_gt_543| and |Pat_simplify_1_325|, where the last
sequence of numbers is inserted by Template Haskell to ensure uniqueness.

@@

\begin{code}
type family Con (c :: Symbol)             :: * -> *
type family Fun (f :: Symbol)             :: * -> *
type family Pat (p :: Symbol) (n :: Nat)  :: * -> *
\end{code}
%
where |Symbol| is the kind of type-level strings in Haskell.
%
Then, our derivation process will synthesize each representation using unique
names, along with a type instance of the corresponding type family.
%
For instance, along with the constructions representations |Con_Text|, |Fun_cat|
and |Pat_simplify_1|, we will automatically derive the following type instances:

@@

% The biggest different appears when deriving the required type instances for the
% |Con|, |Fun| and |Pat| synonyms.
%
However, instead of carrying this type parameter in our generation
specifications, we can avoid it by hiding it behind an existential type:

@@

\begin{code}
genIntTree :: BGen (Tree Int)
genIntTree = genRep at (Tree_Spec apply' Int)
\end{code}
%
Where |apply| is a type family that traverses our generation specification,
applying the |Int| type to each occurrence of |Some|:
%
\looseness=-1

@@

Along with our automated constructions derivation mechanism, the machinery
introduced in this section allows us to specify random generators using a simple
and flexible type-level specification language.
%
The next section evaluates our approach in terms of performance using a set of
case studies extracted from real-world Haskell implementations.
